Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-03-10 man/ecosystems/ms_ecosystems.Rmd
last changes

## Introduction
The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance, higher vulnerability, and thus higher risk of extinction than do populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine under what environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].

@@

The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution. In these cases, climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005]. The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010]. Also, it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].

@@

### Field sampling and dendrochronological methods to assess individual tree resilience
Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1). For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the sampled aspect. Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled. Two cores of 5 mm in diameter were taken from each tree at breast height (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH, and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.

@@

### Statistical analysis 
Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015]. For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used because they were close to the mean value in all cases [@Wilcox2012]. When the robust ANOVA test was run, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers. *Post-hoc* differences were assessed pairwise using a similar bootstrap test. All the robust ANOVA and *post-hoc* tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons.

@@

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly faster than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3). The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 3 and S3a). Only trees from the southern sites (*i.e.* the wetter aspect) showed significant positive growth trends since the late 1970s (Figure 3), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a), this negative effect being particularly evident for the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies. Tree-growth reduction followed a moisture gradient. The southern sites (CA-High and CA-Low) showed less tree-growth reduction than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site.

>>>>

## Introduction
The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016]. It has been assumed that geographical and ecological rear-edge populations match. This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine under what environmental conditions the geographical rear edge does not correspond with the ecological one [@VilaCabreraJump2019GreaterGrowth].

@@

The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution. In these cases, climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005]. The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010]. Also, it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems where land-use legacies have strongly modified species distribution [@VilaCabrera2019]. Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].

@@

### Field sampling and dendrochronological methods to assess individual tree resilience
Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1). For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the site sampled. Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled. Two cores of 5 mm in diameter were taken from each tree at breast height (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH, and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.

@@

### Statistical analysis 
Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015]. For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern exposures). These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used because they were close to the mean value in all cases [@Wilcox2012]. When the robust ANOVA test was run, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers. *Post-hoc* differences were assessed pairwise using a similar bootstrap test. All the robust ANOVA and *post-hoc* tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons.

@@

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly faster than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3). The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 3 and S3a). Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 3), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a). The strongest reduction in radial growth occurred in the 1995 drought (the worst drought spell in our climatic record, Table S3) for all sites. Tree-growth reduction followed a moisture gradient.  The southern sites (CA-High and CA-Low) showed less tree-growth reduction than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site (CA-High).
